---
title: "Simulations preferring distance based"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

### source

```{r}

library(AntMAN)
library(mclust)
library(dplyr)
library(tidyr)
library(ggplot2)
library(salso)
library(mclust)

source("src/mixture_utils.R", echo=TRUE)
source("src/distances.R", echo=TRUE)
source("src/search_algorithm.R", echo=TRUE)

```

This file serves to explore cases in which the density based methods have an advantage over classical allocation based methods like the binder or VI.

# Overlap

There is some reason to believe that overlapping clusters could be easier to find with a density based method. If using mcmc, salso could get worse from label switching.

## Simulate overlapping data

```{r}

set.seed(1)

N = 100
K = 3
alpha = 1
mu_true = c(-2, 0, 1)    
sd_true = c(1, 2, 1) * 0.2   

#--- Component weights from Dirichlet(alpha,...,alpha) ---
weights_true <- c(0.3, 0.5, 0.2)

#--- Assign clusters ---
cluster_true <- rep(c(1:K), weights_true * N)

#--- Initialize data ---
y <- numeric(N)

#--- Generate Gaussian data ---
for (k in 1:K) {
  
  idx <- cluster_true == k
  ni  <- sum(idx)
  y[idx] <- rnorm(ni, mean = mu_true[k], sd = sd_true[k])
  
}

plot(density(y))
points(y, rep(0, N), col = cluster_true)

```

## Run Gaussian mixture (AntMAN)

```{r}

  alpha_0 <- 1
  kappa_0 <- 0.01
  nu_0 <- 3 # alpha of the invgamma prior
  sig2_0 <- 2 # beta of the inverse-gamma prior
  mu_0 <- 0 # prior mean of the mean
  
  # -----------------------------
  # MCMC settings
  # -----------------------------
  n_iter <- 2e4
  burn_in <- n_iter / 4
  thin <- 10
  
  # -----------------------------
  # Define ANTMAN priors & parameters
  # -----------------------------
  mixture_uvn_params <- AntMAN::AM_mix_hyperparams_uninorm(
    m0 = mu_0, k0 = kappa_0, nu0 = nu_0, sig02 = sig2_0
  )
  
  mcmc_params <- AntMAN::AM_mcmc_parameters(
    niter = n_iter, burnin = burn_in, thin = thin, verbose = 1
  )
  
  components_prior <- AntMAN::AM_mix_components_prior_pois(Lambda = K)
  
  # -----------------------------
  # Run MCMC and cluster
  # -----------------------------
  mix_post_draws <- AntMAN::AM_mcmc_fit(
    y = y,
    mix_kernel_hyperparams = mixture_uvn_params,
    mix_components_prior = components_prior,
    mcmc_parameters = mcmc_params
  )
  
  eam <- AM_clustering(mix_post_draws)  # M*n matrix
  
  clustering_matrix <- eam + 1

  binder_loss <- apply(
    clustering_matrix,
    MARGIN = 1,
    FUN = function(cl) binder(truth = cluster_true , estimate = cl)
  )
  
  vi_loss <- apply(
    clustering_matrix,
    MARGIN = 1,
    FUN = function(cl) VI(truth = cluster_true , estimate = cl)
  )
  
```


```{r}

options(error = NULL)
# =============================================================================
# New simulation study using the full search algorithm (initialisation + sweetening)
# (merge-split phase turned off for now – n_ms = 0)
# =============================================================================

prior_list <- list(
  alpha_0 = 1,
  kappa_0 = 0.01,
  mu_0    = 0,
  a_0     = 2,
  b_0     = 5
)

# ----------------------------------------------------------------------------
# Configurations we want to test
# ----------------------------------------------------------------------------
configs <- rbind(
  data.frame(method = "KS",       clusterwise = FALSE, version = "grid",    stringsAsFactors = FALSE),
  data.frame(method = "KS",       clusterwise = TRUE,  version = "grid",    stringsAsFactors = FALSE),
  data.frame(method = "Pearson",  clusterwise = FALSE, version = "grid",    stringsAsFactors = FALSE),
  data.frame(method = "Pearson",  clusterwise = TRUE,  version = "grid",    stringsAsFactors = FALSE),
  data.frame(method = "Wasserstein", clusterwise = FALSE, version = "grid",    stringsAsFactors = FALSE),
  #data.frame(method = "Wasserstein", clusterwise = TRUE,  version = "grid",    stringsAsFactors = FALSE),
  #data.frame(method = "Wasserstein", clusterwise = FALSE, version = "quantile", stringsAsFactors = FALSE),
  data.frame(method = "Wasserstein", clusterwise = TRUE,  version = "quantile", stringsAsFactors = FALSE)
)

# ----------------------------------------------------------------------------
# Run search algorithm for every configuration
# ----------------------------------------------------------------------------
results_list <- vector("list", nrow(configs))

for (i in seq_len(nrow(configs))) {
#i = 1
  cfg <- configs[i, ]
  
  cat(sprintf("Running %s (clusterwise=%s, version=%s)...\n",
              cfg$method, cfg$clusterwise, cfg$version))
  set.seed(1)
  fit <- run_clustering(
    data              = y,
    clustering_matrix = clustering_matrix,
    posterior_draws   = mix_post_draws,
    prior_list        = prior_list,
    method            = cfg$method,
    clusterwise       = cfg$clusterwise,
    estimator         = "mle",
    n_runs            = 2,          # modest number of random starts
    n_sweet           = 3, #80,         # enough for sweetening to converge in this example
    tol               = 1e-10,
    n_ms              = 0,          # ← merge-split turned off (as requested)
    n_merge           = 0,
    n_split           = 0,
    version           = cfg$version,
    M                 = 2000,
    alpha_0           = 5
  )
  # check if there are gaps in the clustering:
  if(all(apply(fit$c_record, 1, max) > 3)){
    if(!any(sapply(apply(fit$c_record, 1, table), length) == apply(fit$c_record, 1, max))){
      stop(paste0("Config ", i, " showed a gap"))
    }
  }else{
    if(!any(dim(apply(fit$c_record, 1, table))[2] == apply(fit$c_record, 1, max))){
      stop(paste0("Config ", i, " showed a gap"))
    }
  }
  
  
  # Take the best run (lowest final distance)
  best_run <- which.min(fit$distance_record)
  
  results_list[[i]] <- data.frame(
    method        = cfg$method,
    clusterwise   = cfg$clusterwise,
    version       = cfg$version,
    final_D       = fit$distance_record[best_run],
    final_c       = I(list(fit$final_clustering)),
    ARI           = adjustedRandIndex(fit$final_clustering, cluster_true),
    n_sweet       = fit$total_accepted_sweets[best_run],
    clustering    = I(list(fit$c_record)),   # keep as list column for later unnest
    stringsAsFactors = FALSE
  )
}

```


```{r}

# ----------------------------------------------------------------------------
# Add Binder and VI as baselines (computed on the raw AntMAN matrix)
# ----------------------------------------------------------------------------
binder_best_idx <- which.min(binder_loss)
vi_best_idx     <- which.min(vi_loss)

results_list[[7]] <- data.frame(
  method      = "Binder",
  clusterwise = FALSE,
  version     = NA_character_,
  final_D     = min(binder_loss),
  final_c       = I(list(clustering_matrix[binder_best_idx, ])),
  ARI         = adjustedRandIndex(clustering_matrix[binder_best_idx, ], cluster_true),
  n_sweet     = NA_integer_,
  clustering  = I(list(clustering_matrix[binder_best_idx, ])),
  stringsAsFactors = FALSE
)

results_list[[8]] <- data.frame(
  method      = "VI",
  clusterwise = FALSE,
  version     = NA_character_,
  final_D     = min(vi_loss),
  final_c       = I(list(clustering_matrix[vi_best_idx, ])),
  ARI         = adjustedRandIndex(clustering_matrix[vi_best_idx, ], cluster_true),
  n_sweet     = NA_integer_,
  clustering  = I(list(clustering_matrix[vi_best_idx, ])),
  stringsAsFactors = FALSE
)



results_df <- do.call(rbind, results_list)

# Create a nice label for faceting/plotting
results_df <- results_df %>%
  mutate(
    label = case_when(
      method %in% c("Binder", "VI") ~ method,
      TRUE ~ paste0(
        method,
        ifelse(clusterwise, " (cw)", ""),
        ifelse(!is.na(version) & method == "Wasserstein", 
               paste0(" | ", version), 
               "")
      )
    )
  )

print(results_df %>% select(label, final_D, ARI, n_sweet))

```

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(mclust)   # for adjustedRandIndex, if not already loaded

# ----------------------------------------------------------------------------
# Prepare truth as a reference row (mimicking the old truth_df)
# ----------------------------------------------------------------------------
truth_row <- tibble(
  label       = "Truth",
  final_D     = NA_real_,
  ARI         = 1.0,
  clustering  = list(cluster_true)
)

# Combine with algorithm results
plot_df <- bind_rows(
  truth_row,
  results_df %>% 
    select(label, final_D, ARI, final_c)%>%
     rename(clustering = final_c)
)  %>% 
  # Add a sorting key so Truth appears first, then ordered by ARI descending
  mutate(
    sort_key = if_else(label == "Truth", Inf, ARI),
    label    = factor(label, levels = unique(label[order(-sort_key)]))
  ) %>%
  arrange(desc(sort_key))

# ----------------------------------------------------------------------------
# Unnest into long format for plotting
# ----------------------------------------------------------------------------
plot_long <- plot_df %>%
  mutate(
    obs_idx = row_number()   # not really needed, but for safety
  ) %>%
  unnest_longer(clustering, indices_to = "point_idx", values_to = "Cluster") %>%
  mutate(
    y      = rep(y, times = nrow(plot_df)),          # repeat data values for each method
    Cluster = factor(Cluster)
  )

# ----------------------------------------------------------------------------
# Prepare ARI annotation positions (placed at right side of each facet)
# ----------------------------------------------------------------------------
labels_df <- plot_df %>%
  mutate(
    label_text = sprintf("ARI = %.3f", ARI),
    x          = max(y) + diff(range(y)) * 0.02,   # slight offset from max y
    y_pos      = 0.1                               # fixed vertical position
  ) %>%
  select(label, label_text, x, y_pos)

# ----------------------------------------------------------------------------
# Plot
# ----------------------------------------------------------------------------
p <- plot_long %>% filter(label != "KS") %>% 
  ggplot(aes(x = y, y = 0, color = Cluster)) +
  geom_point(size = 2.8, alpha = 0.75, stroke = 0.4) +
  geom_text(
    data = labels_df,
    aes(x = x, y = y_pos, label = label_text),
    inherit.aes = FALSE,
    hjust = 1,
    vjust = 0.5,
    size  = 3.8,
    fontface = "bold",
    color = "black"
  ) +
  facet_wrap(~ label, ncol = 1, strip.position = "left") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.title.y      = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    strip.text        = element_text(size = 11, face = "bold", hjust = 0),
    strip.background  = element_rect(fill = "grey92"),
    strip.placement   = "outside",
    legend.position   = "none",               # optional: remove if too many clusters
    plot.title        = element_text(hjust = 0.5, face = "bold"),
    plot.margin       = margin(10, 25, 10, 10)
  ) +
  labs(
    x     = expression(italic(y)),
    title = "Best clustering per method / loss (including ground truth)"
  )

print(p)

# Optional: save with good resolution for paper
# ggsave("best_clusterings_comparison.pdf", p, width = 8, height = 1.4 * nlevels(plot_long$label), dpi = 400)

```


# Conclusion for the sweetening

It seems the sweetening is now successful in a couple of cases which allow the algorithm to estimate clusterings not visited by the MCMC.
By controlling the alpha_0/shape hyperparameter of the variance prior, we can incentivize the creation of new clusters.


Next step is to test the merge and split part


# Merge and Split Test

```{r}
#options(error = recover)
#options(error = NULL)

# =============================================================================
# New simulation study using the full search algorithm (initialisation + sweetening)
# (merge-split phase turned off for now – n_ms = 0)
# =============================================================================

prior_list <- list(
  alpha_0 = 1,
  kappa_0 = 0.01,
  mu_0    = 0,
  a_0     = 2,
  b_0     = 5
)

# ----------------------------------------------------------------------------
# Configurations we want to test
# ----------------------------------------------------------------------------

n_ms = 5
n_merge = 10

configs_ms <- rbind(
  data.frame(method="KS", clusterwise=FALSE, version="grid", n_sweet=1, n_ms = n_ms, n_merge=n_merge, n_split=4),
  data.frame(method="Pearson", clusterwise=FALSE, version="grid", n_sweet=1, n_ms = n_ms, n_merge=n_merge, n_split=4),
  data.frame(method="Wasserstein", clusterwise=TRUE, version="quantile", n_sweet=1, n_ms = n_ms, n_merge=n_merge, n_split=4),
  data.frame(method="Wasserstein", clusterwise=FALSE, version="grid", n_sweet=1, n_ms = n_ms, n_merge=n_merge, n_split=4)
)

# ----------------------------------------------------------------------------
# Run search algorithm for every configuration
# ----------------------------------------------------------------------------
results_list <- vector("list", nrow(configs_ms))

for (i in seq_len(nrow(configs_ms))) {
#i = 1
  cfg <- configs_ms[i, ]
  
  cat(sprintf("Running %s (clusterwise=%s, version=%s)...\n",
              cfg$method, cfg$clusterwise, cfg$version))
  set.seed(1)
  fit <- run_clustering(
    data              = y,
    clustering_matrix = clustering_matrix,
    posterior_draws   = mix_post_draws,
    prior_list        = prior_list,
    method            = cfg$method,
    clusterwise       = cfg$clusterwise,
    estimator         = "mle",
    n_runs            = 3,          # modest number of random starts
    tol               = 1e-10,
    n_sweet   = cfg$n_sweet,
    n_ms      = cfg$n_ms,
    n_merge   = cfg$n_merge,
    n_split   = cfg$n_split,
    version           = cfg$version,
    M                 = 2000,
    alpha_0           = 5
  )
  # check if there are gaps in the clustering:

  has_gap <- apply(fit$c_record, 1, function(row) length(unique(row))) < apply(fit$c_record, 1, max)
  
  if (any(has_gap)) {
    cat(sprintf("Config %d shows labeling gaps in %d rows\n", i, sum(has_gap)))
    print(which(has_gap))
    stop(sprintf("Config %d showed a gap", i))
  }
  
  # After running fit <- run_clustering(...)

best_run <- which.min(fit$distance_record)

n_merge_accept_best <- fit$total_accepted_merges[best_run]
n_split_accept_best <- fit$total_accepted_splits[best_run]


results_list[[i]] <- data.frame(
  method          = cfg$method,
  clusterwise     = cfg$clusterwise,
  version         = cfg$version,
  n_sweet         = cfg$n_sweet,                # requested value
  n_ms            = cfg$n_ms,
  n_merge         = cfg$n_merge,
  n_split         = cfg$n_split,
  final_D         = fit$distance_record[best_run],
  ARI             = adjustedRandIndex(fit$final_clustering, cluster_true),
  n_merge_accept  = n_merge_accept_best,
  n_split_accept  = n_split_accept_best,
  n_sweet_actual  = fit$total_accepted_sweets[best_run],  # actual iterations run
  clustering      = I(list(fit$final_clustering)),        # final assignment vector
  # Optional: keep full history if memory allows and needed for diagnostics
  # c_record        = I(list(fit$c_record)),
  stringsAsFactors = FALSE
)

}




```



```{r}

# ----------------------------------------------------------------------------
# Add Binder and VI as baselines (computed directly on posterior samples)
# ----------------------------------------------------------------------------
binder_best_idx <- which.min(binder_loss)
vi_best_idx     <- which.min(vi_loss)

results_list[[length(results_list) + 1]] <- data.frame(
  method          = "Binder",
  clusterwise     = FALSE,
  version         = NA_character_,
  n_sweet         = NA_integer_,
  n_ms            = NA_integer_,
  n_merge         = NA_integer_,
  n_split         = NA_integer_,
  final_D         = min(binder_loss, na.rm = TRUE),
  ARI             = adjustedRandIndex(clustering_matrix[binder_best_idx, ], cluster_true),
  n_merge_accept  = NA_integer_,
  n_split_accept  = NA_integer_,
  n_sweet_actual  = NA_integer_,
  clustering      = I(list(clustering_matrix[binder_best_idx, ])),
  stringsAsFactors = FALSE
)

results_list[[length(results_list) + 1]] <- data.frame(
  method          = "VI",
  clusterwise     = FALSE,
  version         = NA_character_,
  n_sweet         = NA_integer_,
  n_ms            = NA_integer_,
  n_merge         = NA_integer_,
  n_split         = NA_integer_,
  final_D         = min(vi_loss, na.rm = TRUE),
  ARI             = adjustedRandIndex(clustering_matrix[vi_best_idx, ], cluster_true),
  n_merge_accept  = NA_integer_,
  n_split_accept  = NA_integer_,
  n_sweet_actual  = NA_integer_,
  clustering      = I(list(clustering_matrix[vi_best_idx, ])),
  stringsAsFactors = FALSE
)

# Combine into data frame
results_df <- do.call(rbind, results_list)

# Create faceting label (unchanged)
results_df <- results_df %>%
  mutate(
    label = case_when(
      method %in% c("Binder", "VI") ~ method,
      
      TRUE ~ paste0(
        method,
        if_else(clusterwise, " (cw)", "", missing = ""),
        if_else(
          !is.na(version) & method == "Wasserstein",
          paste0(" | ", version),
          "",
          missing = ""
        )
      )
    ),
    # Optional: nicer sorting — put Truth first if you add it later, then Binder/VI, then methods by ARI or alphabetically
    label = forcats::fct_relevel(label, "Truth", after = 0L),  # if you have "Truth"
    # or just leave as is / use reorder() later for plotting
    label = factor(label)
  )

# Quick summary print
print(results_df %>% 
        select(label, final_D, ARI, n_sweet, n_ms, n_merge_accept, n_split_accept) %>%
        arrange(final_D))

```




