---
title: "Simulation study and sensitivity analysis of the inidvidual parts in 1D"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

# Setup, load and source

```{r setup}
# Load necessary libraries
library(AntMAN)
library(salso)
library(tidyverse)
library(Rcpp)
library(RcppArmadillo)
library(fossil)
library(dplyr)
library(mclust)  # for adjustedRandIndex

library(tidyr)
library(ggplot2)

getwd()
source("src/mixture_utils.R")

```

### general setup

```{r}

n_replication <- 1

for(i in 1:n_replication){
  
  set.seed(i)
  
  N <- 200
  K <- 3
  alpha_0 <- 1
  
  simdata <- generate_mixture_data(
      N,
      K,
      dim = 1,
      alpha = alpha_0,
      mu_true = NULL,      
      Sigma_true = NULL 
  )
  
  y <- simdata$data
  cluster_true <- simdata$cluster_true
  weights_true <- simdata$weights_true
  
  # -----------------------------
  # Prior hyperparameters
  # -----------------------------
  # invgamma mean is beta / alpha -1
  # to accomodate mean values of -K to K, lets say for K = 3 around a prior mean of the mean of 0 we need
  # an sd close to K. Therefore b / a - 1 = K. The variance is  
  
  alpha_0 <- 1
  kappa_0 <- 0.1
  nu_0 <- 3 # alpha of the invgamma prior
  sig2_0 <- 2 # beta of the inverse-gamma prior
  mu_0 <- 0 # prior mean of the mean
  
  
  # Constellation from XJs simulation
  # alpha = 1
  # kappa_0 = 0.01
  # mu_0 = 0
  # nu_0 = 2
  # sig2_0 = 5
  
  prior_list <- list(
    alpha_0 = 1,       # Dirichlet concentration for mixture weights
    kappa_0 = 1,   
    mu_0    = 0,       # prior mean for component means
    a_0     = 2,       # shape parameter for inverse-gamma prior on variance
    b_0     = 1        # scale parameter for inverse-gamma prior on variance
  )
  # -----------------------------
  # MCMC settings
  # -----------------------------
  n_iter <- 2e4
  burn_in <- n_iter / 4
  thin <- 10
  
  # -----------------------------
  # Define ANTMAN priors & parameters
  # -----------------------------
  mixture_uvn_params <- AntMAN::AM_mix_hyperparams_uninorm(
    m0 = mu_0, k0 = kappa_0, nu0 = nu_0, sig02 = sig2_0
  )
  
  mcmc_params <- AntMAN::AM_mcmc_parameters(
    niter = n_iter, burnin = burn_in, thin = thin, verbose = 1
  )
  
  components_prior <- AntMAN::AM_mix_components_prior_pois(Lambda = K)
  
  # -----------------------------
  # Run MCMC and cluster
  # -----------------------------
  mix_post_draws <- AntMAN::AM_mcmc_fit(
    y = y,
    mix_kernel_hyperparams = mixture_uvn_params,
    mix_components_prior = components_prior,
    mcmc_parameters = mcmc_params
  )
  
  eam <- AM_clustering(mix_post_draws)  # M*n matrix
  
  clustering_matrix <- eam + 1
  
  res_KS <- run_clustering(data = y,
                 clustering_matrix = clustering_matrix,
                 posterior_draws = mix_post_draws,
                 prior_list,
                 method         = "KS",
                 n_runs         = 1,
                 n_sweet = 0,
                 tol            = 1e-10,
                 n_ms          = 10,
                 n_merge        = 10,
                 n_split        = 0)
    
    
  
}

# plot(y, rep(0, N), col = cluster_true)
# 
# plot(y, rep(0, N), col = clustering_matrix[200,])
# plot(y, rep(0, N), col = clustering_matrix[1000,])
plot(y, rep(0, N), col = res_KS$clustering)

# iter <- 200
# cols <- clustering_matrix[iter, ]
# ucl  <- paste0("cluster:", sort(unique(cols))," mean:", round(unlist(mix_post_draws$mu[[iter]]), 2), " var:", round(unlist(mix_post_draws$sig2[[iter]]), 2))
# 
# plot(y, rep(0, N), col = cols)
# legend("topright",
#        legend = ucl,
#        col = sort(unique(cols)),
#        pch = 1,
#        title = "Cluster")


```

### Sensitivity of the distances to the number of clusters

Quick comparison if the distance metrics prefer less or more clusters. Start by analyzing the binder results:

#### Binder

```{r}


binder_loss <- apply(
  clustering_matrix,
  MARGIN = 1,
  FUN = function(cl) binder(truth = cluster_true , estimate = cl)
)

par(mfrow = c(2, 1))
plot(apply(clustering_matrix, 1, max), type = "l")
plot(binder_loss, col = "red", type = "l")

c_binder <- clustering_matrix[which.min(binder_loss),]

```

```{r}

Kmax <- apply(clustering_matrix, 1, max)
binder_summary <- aggregate(
  binder_loss,
  by = list(Kmax = Kmax),
  FUN = function(x) c(mean = mean(x), var = var(x))
)

```

Seems that the binder correctly favors 3 clusters and punishes solutions with more clusters.

#### Distances

What about our distances:

```{r}

methods <- c("KS", "Pearson", "Wasserstein")
n_candidates <- nrow(clustering_matrix)

# Precompute max cluster label for each clustering
Kmax <- apply(clustering_matrix, 1, max)

# Initialize a list to store distance vectors per method
distance_list <- vector("list", length(methods))
distance_list_cw <- vector("list", length(methods))
names(distance_list) <- methods

prior_list <- list(
  alpha_0 = 1,       # Dirichlet concentration for mixture weights
  kappa_0 = 1,   
  mu_0    = 0,       # prior mean for component means
  a_0     = 2,       # shape parameter for inverse-gamma prior on variance
  b_0     = 1        # scale parameter for inverse-gamma prior on variance
)

# XJ prior list:
prior_list <- list(
  alpha_0 = 1,       # Dirichlet concentration for mixture weights
  kappa_0 = 0.01,   
  mu_0    = 0,       # prior mean for component means
  a_0     = 2,       # shape parameter for inverse-gamma prior on variance
  b_0     = 5        # scale parameter for inverse-gamma prior on variance
)

for (method in methods) {
  distance <- numeric(n_candidates)
  distance_cw <- numeric(n_candidates)
  
  for (i in seq_len(n_candidates)) {
    c_i <- clustering_matrix[i, ]
    param_post <- mixture_posterior(c_i, y, prior_list)
    w_i <- param_post$weight_post
    mu_i <- param_post$mu_post
    var_i <- param_post$var_post
    
    distance[i] <- compute_distance(data, w_i, mu_i, var_i, method)
    distance_cw[i] <- compute_distance(data, w_i, mu_i, var_i, method, clusterwise = TRUE, c_alloc = c_i)
  }
  
  distance_list[[method]] <- distance
  distance_list_cw[[method]] <- distance_cw
}

# Combine results into a data.frame for summarizing
results_df <- do.call(rbind, lapply(methods, function(m) {
  data.frame(
    method = m,
    Kmax = Kmax,
    distance = distance_list[[m]],
    cw = FALSE
  )
}))

results_df_cw <- do.call(rbind, lapply(methods, function(m) {
  data.frame(
    method = m,
    Kmax = Kmax,
    distance = distance_list_cw[[m]],
    cw = TRUE
  )
}))

results_df <- rbind(results_df, results_df_cw)


```


```{r}

summary_df <- results_df %>%
  group_by(method, Kmax, cw) %>%
  summarise(
    mean_distance = mean(distance),
    var_distance = var(distance),
    .groups = "drop"
  )


summary_df %>% print(n = 21)

summary_df %>% 
  #filter(method != "Wasserstein") %>% 
ggplot(aes(x = factor(Kmax), y = mean_distance, fill = factor(cw))) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_errorbar(aes(
    ymin = mean_distance - sqrt(var_distance),
    ymax = mean_distance + sqrt(var_distance)
  ), 
  width = 0.2,
  position = position_dodge(0.9)) +
  facet_wrap(~method, scales = "free") + 
  theme_minimal()

```

Old:
It seems the distances do not pick up the right solutions, Wasserstein and KS even prefer a higher number of clusters! That would make sense since the density should be approximated better with a higher number of clusters.
New: after changing the priors, the distances seem to work!


Let's see if the preferred solutions are accurate, irrespective of the number of clusters:

```{r}


# Assume:
# clustering_matrix: matrix with clusterings (rows = clusterings, cols = data points)
# results_df: data frame with columns method, distance, and (added below) index
# y: numeric vector (data)
# cluster_true: true cluster labels vector (factor or integer)

n_candidates <- nrow(clustering_matrix)
results_df$index <- rep(1:n_candidates, times = length(unique(results_df$method)))

# Step 1: Find best clustering index per method
best_solutions <- results_df %>%
  group_by(method, cw) %>%
  slice_min(distance, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(method, cw, best_index = index)


# Step 2: Extract best clusterings
best_clusterings <- lapply(best_solutions$best_index, function(idx) {
  clustering_matrix[idx, ]
})

best_clusterings[[7]] <- c_binder

names(best_clusterings) <- c(paste(
  best_solutions$method,
  best_solutions$cw,
  sep = " | "
), "binder")


#Step 3: Prepare plotting data frame
plot_df <- data.frame(
  y = y,
  true = factor(cluster_true)
)

for (nm in names(best_clusterings)) {
  plot_df[[nm]] <- factor(best_clusterings[[nm]])
}

plot_long <- plot_df %>%
  pivot_longer(
    cols = -y,
    names_to = "Clustering",
    values_to = "Cluster"
  )

ggplot(plot_long, aes(x = y, y = 0, color = Cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  facet_wrap(~ Clustering, ncol = 1) +
  theme_minimal() +
  theme(
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  labs(
    x = "Data (y)",
    y = "",
    color = "Cluster",
    title = "Comparison of Best Clusterings per Method and cw vs True Labels"
  )




```


The distance methods are very sensitive to the prior choice!!

# Adjusted Rand Index

```{r}



# Compute Rand Index and Adjusted Rand Index for each clustering
results <- lapply(best_clusterings, function(clust) {
  ari <- adjustedRandIndex(clust, cluster_true)
})

# Convert results to a data.frame for better readability
results_df <- do.call(rbind, lapply(names(results), function(name) {
  data.frame(
    Solution = name,
    Adjusted_Rand_Index = results[[name]]
  )
}))

print(results_df)

```

```{r}


# Rename 'Solution' to 'Clustering' to join with plot_long
results_df2 <- results_df %>% 
  rename(Clustering = Solution)

# Join ARI to plot_long (just for completeness, though not strictly necessary for plotting)
plot_long2 <- plot_long %>%
  left_join(results_df2, by = "Clustering")

# Create labels for ARI with rounded values
labels_df <- results_df2 %>%
  mutate(label_text = paste0("ARI: ", round(Adjusted_Rand_Index, 3)))

# Get x and y positions to place text (e.g., max x and y=0.1)
text_positions <- plot_long %>%
  group_by(Clustering) %>%
  summarise(x_pos = max(y), y_pos = 0.1, .groups = "drop")

labels_df <- labels_df %>%
  left_join(text_positions, by = "Clustering")

# Plot with ARI annotations
ggplot(plot_long2, aes(x = y, y = 0, color = Cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(
    data = labels_df,
    aes(x = x_pos, y = y_pos, label = label_text),
    inherit.aes = FALSE,
    hjust = 1,
    #vjust = 0.1,
    size = 4,
    color = "black"
  ) +
  facet_wrap(~ Clustering, ncol = 1) +
  theme_minimal() +
  theme(
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  labs(
    x = "Data (y)",
    y = "",
    color = "Cluster",
    title = "Comparison of Best Clusterings per Method and cw vs True Labels"
  )

```




### XJs method:

KS:

```{r}

Mixture_normal_cdf = function(x,w,u,s) sum( w*pnorm(x,mean=u,sd=s) )

mixture_posterior <- function(c_alloc, data, prior_list){
  # prior
  alpha = prior_list$alpha
  kappa0 = prior_list$kappa0
  theta = prior_list$theta
  a = prior_list$a
  b = prior_list$b
  
  
  cluster=sort(unique(c_alloc))
  K = length(cluster)
  N = length(data)
  data_ordered = sort(data)
  
  alpha_post = numeric(K)
  mu_post = numeric(K)
  sigma2_post = numeric(K)
  
  # posterior mean
  for (k in 1:K) {
    I = c_alloc==cluster[k]
    nk=sum(I)
    yk=data[I]
    alpha_post[k]=alpha+sum(c_alloc==cluster[k])
    kappa_post = kappa0 +nk
    mu_post[k] = (kappa0 * theta + sum(yk))/kappa_post
    a_post = a + (nk + 1) * 0.5 # if mean depends on sigma^2, need to add 1!
    b_post = b + 0.5 * sum((yk-mean(yk))^2) + 0.5*kappa0*nk*(mean(yk)-theta)^2/kappa_post
    sigma2_post[k] = b_post/(a_post -1) 
  }
  weights= alpha_post/sum(alpha_post)   
  
  posterior_list = list("weights"=weights, "mu_post"=mu_post, "sigma2_post"=sigma2_post)
  return(posterior_list)
}


KS_distance_mixture <- function(observed_data, c_alloc, prior_list){
  
  posterior_list = mixture_posterior(c_alloc, observed_data, prior_list)
  w = posterior_list$weights
  u = posterior_list$mu_post
  s = posterior_list$sigma2_post
  
  ecdf_func <- function(data) { 
    Length <- length(data) 
    sorted <- sort(data) 
    ecdf <- rep(0, Length) 
    for (i in 1:n) { 
      ecdf[i] <- sum(sorted <= data[i]) / Length 
    } 
    return(ecdf) 
  } 
  
  n = length(observed_data)
  FN = ecdf_func(observed_data)
  FN_sort = sort(FN)
  knots_vec = sort(observed_data)
  distance_vec = numeric(n-1)
  distance_vec[1] = abs(Mixture_normal_cdf(knots_vec[1], w, u, s) - FN_sort[1])
  for(i in 2:n){
    x_i = knots_vec[i-1]
    x_i_next = knots_vec[i]
    d1 = abs(Mixture_normal_cdf(x_i,w, u, s) - FN_sort[i])
    d2 = abs(Mixture_normal_cdf(x_i_next,w, u, s) - FN_sort[i])
    distance_vec[i] = max(d1, d2)
  }
  KS = max(distance_vec)
  return(KS)
}





```

```{r}

Initialization_phase <- function(c_sample, observed_data, prior_list, distance_function){
  # Input: mat I_mcmc*n c_sample
  #        double [0,1] p_sa: the probability to sequential allocation
  #        int k_d: the number of maximum clusters
  # print("Start Initialization Phase")
  result = min_distance(c_sample, observed_data, prior_list,distance_function) + 1
  # print("End Initialization Phase")
  return(result)
}

min_distance <- function(c_samples, data, prior_list, distance_function){
  # c_samples: M*n matrix
  M = dim(c_samples)[1]
  result_vec = numeric(M)
  for (i in 1:M) {
    if(i%%100==0){
      print(i)
    }
    result_vec[i] = distance_function(data, c_samples[i,], prior_list)
  }
  min_m = which(result_vec==min(result_vec))[1]
  # print(min_m)
  return(c_samples[min_m,])
}

```

```{r}

alpha = 1
kappa0 = 0.01
theta = 0
a = 2
b = 5
prior_list = list("alpha"=alpha, "kappa0"=kappa0, "theta"=theta, "a"=a, "b"=b)
data = y
c_samples = clustering_matrix

M = dim(c_samples)[1]
result_vec = numeric(M)
for (i in 1:M) {
  if(i%%100==0){
    print(i)
  }
  result_vec[i] = KS_distance_mixture(data, c_samples[i,], prior_list)
}
min_m = which(result_vec==min(result_vec))[1]



```

Add it to the rest:

```{r}

#results_df_all <- cbind(rbind(results_df, data.frame(method = "KS_XJ", Kmax = 99, distance = result_vec, index = 1:M)), binder_loss =rep(binder_loss, 4))

results_df_all <- cbind(results_df, binder_loss =rep(binder_loss, 3))

results_df_all %>% 
ggplot(aes(x = binder_loss, y = distance, col = cw)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ method, scales = "free", ncol = 2) +
  labs(
    x = "Binder Loss",
    y = "Distance",
    title = "Scatterplot of Distance vs Binder Loss by Method"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

```

We can see that except for the Pearson distance, the density based distances have a positive correlation with the binder loss and now give sensible results (see point plot).

We can also see that the cw methods always result in a higher distance than their non-cw alternatives. Would about their accuracy?

```{r}

wide_df <- results_df_all %>%
  select(method, index, cw, distance) %>%  # replace `value` with actual column to compare
  pivot_wider(names_from = cw, values_from = distance,names_prefix = "cw_") 



ggplot(wide_df, aes(x = cw_FALSE, y = cw_TRUE)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ method, scales = "free", ncol = 1) +
  labs(
    x = "Non CW",
    y = "CW",
    title = "Scatterplot of CW vs Non CW by Method"
  ) +
  expand_limits(x = 0, y = 0) +  # ensure axes include 0
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))


```


