---
title: "Sliced Wasserstein sensitivity analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

### source

```{r}

library(mclust)
library(dplyr)
library(tidyr)
library(ggplot2)
source("C:/Users/alexm/NUS Dropbox/Alexander Mozdzen/distance_clustering/src/mixture_utils.R", echo=TRUE)

```

### Simulate Data

```{r}

# -----------------------------
# Simulation parameters
# -----------------------------
N <- 100
K <- 3
D = 2
alpha_0 <- 1

set.seed(2)
  
simdata <- generate_mixture_data(
    N,
    K,
    dim = D,
    alpha = alpha_0,
    mu_true = NULL,      
    Sigma_true = NULL 
)

y <- simdata$data
cluster_true <- simdata$cluster_true
weights_true <- simdata$weights_true
Sigma_true <- simdata$Sigma_true

plot(y,  col = cluster_true)

```

Idea: test the accuracy of the sliced wasserstein distance in comparing the "posterior" parameters corresponding to more or less accurate clusterings.

### 3 different scenarios

We compare the SW for the true clustering, a completely wrong clustering and a reasonable estimate.

```{r}

c_mid <- cluster_true
c_mid[c_mid == 2] <- 1
c_mid[c_mid == 3] <- 2
c_wrong <- cluster_true[sample(100)]

par(mfrow=c(1,3))
plot(y,  col = cluster_true)
plot(y,  col = c_mid)
plot(y,  col = c_wrong)


```

## Posterior estimate conditional on the different clusterings using the initialization phase of our algorithm

```{r}

compute_distance <- function(data, weight, mean, var, method = "KS", clusterwise = FALSE,
                             c_alloc = NULL, L = 100) {
  
  if (is.null(dim(data))) {
    D <- 1
  } else {
    D <- ncol(data)
  }
  
  if(D != 1 & method != "Wasserstein"){
    stop("only Sliced Wasserstein for higher dimensions")
  }
  
  K = length(weight)
  
  if (clusterwise) {
    
    if (is.null(c_alloc)) {
      stop("clusterwise = TRUE requires c_alloc")
    }
    
    if (!is.null(dim(data))) {
      stop("cluster-wise distance currently only implemented for 1D")
    }
    
    K <- length(weight)
    Dk <- numeric(K)
    
    for (k in seq_len(K)) {
      yk <- data[c_alloc == k]
      
      # skip empty / degenerate clusters
      if (length(yk) < 2) {
        Dk[k] <- 0
        next
      }
      
      # component-wise distance
      if (method == "KS") {
        Dk[k] <- KS_mixture_distance(
          yk,
          weight = 1,
          mean   = mean[k],
          sig2   = var[k]
        )
        
      } else if (method == "Pearson") {
        Dk[k] <- Pearson_mixture_distance(
          yk,
          weight = 1,
          mean   = mean[k],
          sig2   = var[k]
        )
        
      } else if (method == "Wasserstein") {
        
        M <- 2e3
        t <- seq(
          from = min(yk) - 3 * sqrt(var[k]),
          to   = max(yk) + 3 * sqrt(var[k]),
          length.out = M
        )
        
        I0 <- mixture_pdf_1D(t, mean[k], sig2 = var[k], weight = 1)
        I1 <- emp_pdf(t, yk)
        
        Dk[k] <- Wasserstein_1D(I0, I1, p = 2)
        
      } else {
        stop("Unknown distance method")
      }
    }
    
    # mixture-weight–weighted aggregation
    return(sum(weight * Dk))
    
  }else{
    
    if (method == "KS") {
      KS_mixture_distance(data, weight, mean, sig2 = var)
    }else if (method == "Wasserstein") {
      # for now always p = 2, consider adapting
      M = 2e3
      t <- seq(
        from = -max(abs(y)) * sqrt(2 * D),
        to   =  max(abs(y)) * sqrt(2 * D),
        length.out = M
      )
      
      if(D == 1){
        
        I0 = mixture_pdf_1D(t, mean, sig2 = var, weight)
        I1 = emp_pdf(t, data)
        Wasserstein_1D(I0, I1, p = 2)
        
      }else{ # Sliced Wasserstein
        
        Sigma = var
        
        #L = 100 # for now fixed
        
        theta = generate_theta_normal(L, D)
        yproj = y %*% t(theta)
        
        projected_Sigma <- matrix(0, nrow = K, ncol = L)
        projected_Mu    <- matrix(0, nrow = K, ncol = L)
        
        for (k in 1:K) {
          for (l in 1:L) {
            projected_Sigma[k, l] <- sqrt( theta[l, , drop = FALSE] %*% Sigma[[k]] %*% t(theta[l, , drop = FALSE]) )
            projected_Mu[k, l]    <- theta[l, , drop = FALSE] %*% mean[[k]]
          }
        }
        
        SW = 0
        
        for (l in 1:L){
          RIx = mixture_pdf_1D(t, projected_Mu[,l], projected_Sigma[,l], weight)
          RIy = emp_pdf(t, yproj[,l])
          SW = SW + Wasserstein_1D(RIx, RIy,p=2) / L
        }
        
        SW
        
      }
      
    } else if (method == "Pearson") {
      Pearson_mixture_distance(data, weight, mean, sig2 = var)
    } else {
      stop("Unknown distance method")
    }
  }
  
}
```


```{r}

D <- ncol(y)

prior_list <- list(
  alpha_0 = 1,
  kappa_0 = 1,
  mu_0    = 0,
  nu_0    = 5,
  S_0     = diag(D)
)

clustering_matrix <- rbind(cluster_true, c_mid, c_wrong)
n_candidates <- nrow(clustering_matrix)

estimators <- c("posterior", "mle")

# storage
distance <- matrix(NA, nrow = n_candidates, ncol = length(estimators),
                   dimnames = list(
                     paste0("cand_", seq_len(n_candidates)),
                     estimators
                   ))

all_params <- vector("list", length(estimators))
names(all_params) <- estimators

for (est in estimators) {
  
  message("Estimator: ", est)
  params_est <- vector("list", n_candidates)
  
  for (i in seq_len(n_candidates)) {
    c_i <- clustering_matrix[i, ]
    
    param <- get_mixture_params(
      c_alloc    = c_i,
      data       = y,
      estimator  = est,
      prior_list = prior_list
    )
    
    params_est[[i]] <- param
    
    distance[i, est] <- compute_distance(
      y,
      weight = param$weight_post,
      mean   = param$mu_post,
      var    = param$var_post,
      method = method
    )
  }
  
  all_params[[est]] <- params_est
}

distance


```

It seems the SW can distinguish correctly between these 3 toy examples and the mle results in lower distances than the posterior.

## Sensitivity wrt the prior parameters

```{r}

## =========================
## Sensitivity analysis over (nu_0, S_0, alpha_0)
## =========================

D <- ncol(y)

# Fixed hyperparameters
mu_0    <- 0
kappa_0 <- 1

# Sensitivity grids
alpha_grid <- c(0.1, 1, 10)                 # mixture weight prior
nu_grid    <- c(D + 1, D + 3, D + 10)        # IW df (must be > D - 1)
S_scale    <- c(0.5, 1, 2)                   # scale multipliers for S_0

# Base scale matrix
S_base <- diag(D)

# Candidate clusterings
clustering_matrix <- rbind(cluster_true, c_mid, c_wrong)
n_candidates <- nrow(clustering_matrix)

# Storage
results <- list()
res_id <- 1

for (alpha_0 in alpha_grid) {
  for (nu_0 in nu_grid) {
    for (s_mult in S_scale) {

      prior_list <- list(
        alpha_0 = alpha_0,
        kappa_0 = kappa_0,
        mu_0    = mu_0,
        nu_0    = nu_0,
        S_0     = S_base * s_mult
      )

      distance <- numeric(n_candidates)
      all_params <- vector("list", n_candidates)

      for (i in seq_len(n_candidates)) {
        c_i <- clustering_matrix[i, ]
        param_post <- mixture_posterior(c_i, y, prior_list)

        w_i   <- param_post$weight_post
        mu_i  <- param_post$mu_post
        var_i <- param_post$var_post

        all_params[[i]] <- param_post
        distance[i] <- compute_distance(y, w_i, mu_i, var_i, method)
      }

      results[[res_id]] <- list(
        alpha_0 = alpha_0,
        nu_0    = nu_0,
        S_mult  = s_mult,
        distance = distance,
        params   = all_params
      )
      res_id <- res_id + 1
    }
  }
}

# Optional: tidy summary table (one row per prior setting and candidate)
summary_df <- do.call(rbind, lapply(results, function(r) {
  data.frame(
    alpha_0 = r$alpha_0,
    nu_0    = r$nu_0,
    S_mult  = r$S_mult,
    candidate = seq_along(r$distance),
    distance  = r$distance
  )
}))

summary_df


```


```{r}

D <- ncol(y)

# Fixed hyperparameters
mu_0    <- 0
kappa_0 <- 1

# Sensitivity grids (posterior only)
alpha_grid <- c(0.1, 1, 10)
nu_grid    <- c(D + 1, D + 3, D + 10)
S_scale    <- c(0.5, 1, 2)

S_base <- diag(D)

# Candidate clusterings
clustering_matrix <- rbind(cluster_true, c_mid, c_wrong)
n_candidates <- nrow(clustering_matrix)

results <- list()
res_id <- 1

## -------------------------
## 1) Posterior sensitivity
## -------------------------
for (alpha_0 in alpha_grid) {
  for (nu_0 in nu_grid) {
    for (s_mult in S_scale) {

      prior_list <- list(
        alpha_0 = alpha_0,
        kappa_0 = kappa_0,
        mu_0    = mu_0,
        nu_0    = nu_0,
        S_0     = S_base * s_mult
      )

      distance <- numeric(n_candidates)
      all_params <- vector("list", n_candidates)

      for (i in seq_len(n_candidates)) {
        c_i <- clustering_matrix[i, ]

        param <- get_mixture_params(
          c_alloc    = c_i,
          data       = y,
          estimator  = "posterior",
          prior_list = prior_list
        )

        distance[i] <- compute_distance(
          y,
          param$weight_post,
          param$mu_post,
          param$var_post,
          method
        )

        all_params[[i]] <- param
      }

      results[[res_id]] <- list(
        estimator = "posterior",
        alpha_0   = alpha_0,
        nu_0      = nu_0,
        S_mult    = s_mult,
        distance  = distance,
        params    = all_params
      )
      res_id <- res_id + 1
    }
  }
}

## -------------------------
## 2) MLE (single baseline)
## -------------------------
distance_mle <- numeric(n_candidates)
params_mle <- vector("list", n_candidates)

for (i in seq_len(n_candidates)) {
  c_i <- clustering_matrix[i, ]

  param <- get_mixture_params(
    c_alloc   = c_i,
    data      = y,
    estimator = "mle"
  )

  distance_mle[i] <- compute_distance(
    y,
    param$weight_post,
    param$mu_post,
    param$var_post,
    method
  )

  params_mle[[i]] <- param
}

results[[res_id]] <- list(
  estimator = "mle",
  alpha_0   = NA,
  nu_0      = NA,
  S_mult    = NA,
  distance  = distance_mle,
  params    = params_mle
)


```

### Visualize

```{r}

summary_df <- do.call(rbind, lapply(results, function(r) {
  data.frame(
    estimator = r$estimator,
    alpha_0   = r$alpha_0,
    nu_0      = r$nu_0,
    S_mult    = r$S_mult,
    candidate = seq_along(r$distance),
    distance  = r$distance
  )
}))
  
summary_df <- summary_df %>%
  mutate(
    estimator = factor(estimator, levels = c("posterior", "mle")),
    alpha_0   = factor(alpha_0),
    nu_0      = factor(nu_0),
    S_mult    = factor(S_mult),
    candidate = factor(candidate)
  )

p1 <- summary_df %>%
  filter(estimator == "posterior") %>%
  ggplot(aes(x = nu_0, y = distance,
             color = S_mult, group = S_mult)) +
  geom_point(alpha = 0.7) +
  geom_line(alpha = 0.6) +
  facet_grid(candidate ~ alpha_0, scales = "free_y") +
  labs(
    title = "Posterior sensitivity to (nu_0, S_0) by candidate",
    x = expression(nu[0]),
    y = "Distance",
    color = expression("Scale of " * S[0])
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

print(p1)

# Clearly shows that a lower scale and higher nu give better results except for a very few cases where the scales 0.5 and 1 change rank with different nu_0
# That would mean that in this case components with a smaller and more concentrated variance are preferred.
# alpha 0.1 and 1 are fine while 10 apparently forces too many clusters and is worse.

#Do all constellations pick up the correct ranking of the toy example?


p2 <- summary_df %>%
  filter(estimator == "posterior") %>%
  ggplot(aes(x = candidate, y = distance,
             color = S_mult, group = S_mult)) +
  geom_point(alpha = 0.7) +
  geom_line(alpha = 0.6) +
  facet_grid(nu_0 ~ alpha_0, scales = "free_y") +
  labs(
    title = "Ranking stability across (nu_0, S_0)",
    x = "Candidate",
    y = "Distance",
    color = expression("Scale of " * S[0])
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

print(p2)

# Yes!

```


### Posterior vs. MLE

Looks a bit messy since faceted by nu and alpha and MLE does not have those but overall clear to see that MLE works better.

```{r}

p3 <- summary_df %>%
  ggplot(aes(x = candidate, y = distance, color = estimator)) +
  geom_point(position = position_dodge(width = 0.4), size = 2) +
  facet_grid(nu_0 ~ alpha_0, scales = "free_y") +
  labs(
    title = "Posterior vs MLE distances",
    x = "Candidate",
    y = "Distance",
    color = "Estimator"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

print(p3)


```



### Sensitivity to L

```{r}

sensitivity_SW_L <- function(
  data,
  weight,
  mean,
  Sigma,
  L_grid = c(10, 20, 50, 100, 200, 500),
  n_rep = 10
) {
  
  results <- expand.grid(
    L = L_grid,
    rep = seq_len(n_rep)
  )
  
  results$SW <- NA_real_
  
  for (i in seq_len(nrow(results))) {
    
    L_i <- results$L[i]
    
    results$SW[i] <- compute_distance(
      data   = data,
      weight = weight,
      mean   = mean,
      var  = Sigma,
      method = "Wasserstein",
      L      = L_i    # see note below
    )
      
  }
  
  results
}


```

```{r}


library(dplyr)

res <- sensitivity_SW_L(
  data   = y,
  weight = w_i,
  mean   = mu_i,
  Sigma  = var_i,
  n_rep  = 10
)

summary_L <- res %>%
  group_by(L) %>%
  summarise(
    SW_mean = mean(SW),
    SW_sd   = sd(SW),
    .groups = "drop")
  
summary_L    
    

```

It seems L = 100 should suffice for our purposes now, though depending on the algorithm we might want to trade accuracy for speed.

# Test on posterior clustering estimates

```{r}
# -----------------------------
# Simulation parameters
# -----------------------------
N <- 100
K <- 3
D = 2
alpha_0 <- 1

set.seed(2)
  
simdata <- generate_mixture_data(
    N,
    K,
    dim = D,
    alpha = alpha_0,
    mu_true = NULL,      
    Sigma_true = NULL 
)

y <- simdata$data
cluster_true <- simdata$cluster_true
weights_true <- simdata$weights_true
Sigma_true <- simdata$Sigma_true

# -----------------------------
# Prior hyperparameters
# -----------------------------
alpha <- 1
kappa_0 <- 0.1
Lam_0 <- diag(D) 
mu_0 <- rep(0, D)
nu_0 <- D + 2

# -----------------------------
# MCMC settings
# -----------------------------
n_iter <- 1e3
burn_in <- n_iter / 4
thin <- 10

# -----------------------------
# Define ANTMAN priors & parameters
# -----------------------------
mixture_mvn_params <- AntMAN::AM_mix_hyperparams_multinorm(
 mu0 = mu_0, ka0 = kappa_0, nu0 = nu_0, Lam0 = Lam_0
)

mcmc_params <- AntMAN::AM_mcmc_parameters(
  niter = n_iter, burnin = burn_in, thin = thin, verbose = 1
)

components_prior <- AntMAN::AM_mix_components_prior_pois(Lambda = K)

# weights_prior <- AntMAN::AM_mix_weights_prior_gamma(init = 2, a = 1, b = 1)

# -----------------------------
# Run MCMC and cluster
# -----------------------------
mix_post_draws <- AntMAN::AM_mcmc_fit(
  y = y,
  mix_kernel_hyperparams = mixture_mvn_params,
  mix_components_prior = components_prior,
  # mix_weight_prior = weights_prior,
  mcmc_parameters = mcmc_params
)

eam <- AM_clustering(mix_post_draws)  # M*n matrix

```

Helper function for the following analysis:

```{r}

compute_distances <- function(
  data,
  clustering_matrix,
  param_method = c("mcmc", "postmean", "mle"),
  prior_list = NULL,
  posterior_params = NULL,
  distance_method = "KS"
) {
  
  param_method <- match.arg(param_method)
  n_candidates <- nrow(clustering_matrix)
  dist <- numeric(n_candidates)
  
  for (i in seq_len(n_candidates)) {
    params <- get_mixture_params(
      clustering       = clustering_matrix[i, ],
      data             = data,
      param_method     = param_method,
      prior_list       = prior_list,
      posterior_params = posterior_params,
      candidate_id     = i
    )
    
    dist[i] <- compute_distance(
      data,
      params$weight,
      params$mean,
      params$var,
      method = distance_method
    )
  }
  
  dist
}



```

```{r}

prior_list <- list(
  alpha_0 = 1,
  kappa_0 = 1,
  mu_0    = 0,
  a_0     = 2,
  b_0     = 1,
  nu_0    = 5,
  S_0     = diag(D)
)

clustering_matrix <- eam + 1
posterior_params <- prepare_AM_posterior_params(mix_post_draws)

dist_mcmc <- compute_distances(
  data = data,
  clustering_matrix = clustering_matrix,
  param_method = "mcmc",
  posterior_params = posterior_params,
  distance_method = "Wasserstein"
)

dist_postmean <- compute_distances(
  data = data,
  clustering_matrix = clustering_matrix,
  param_method = "postmean",
  prior_list = prior_list,
  distance_method = "Wasserstein"
)

dist_mle <- compute_distances(
  data = data,
  clustering_matrix = clustering_matrix,
  param_method = "mle",
  distance_method = "Wasserstein"
)

results_df <- data.frame(
  candidate     = seq_len(nrow(clustering_matrix)),
  dist_mcmc     = dist_mcmc,
  dist_postmean = dist_postmean,
  dist_mle      = dist_mle
)

print(results_df)



```


Compute the corresponding binder estimates:

```{r}

binder_ests <- rep(NA, n_candidates)
vi_ests <- rep(NA, n_candidates)

for(i in nrow(eam)) {
binder_ests[i] <- salso::binder(cluster_true,  clustering_matrix[i, ])
vi_ests[i] <- salso::VI(cluster_true,  clustering_matrix[i, ])
}

results_df$binder_ests <- binder_ests 

```


```{r}

library(mclust)
library(dplyr)
library(ggplot2)

# --- Select best clusterings ---
best_ids <- c(
  Truth     = NA,
  MCMC      = which.min(results_df$dist_mcmc),
  PostMean  = which.min(results_df$dist_postmean),
  MLE       = which.min(results_df$dist_mle),
  Binder    = which.min(results_df$binder_ests)
)

best_clusterings <- list(
  Truth    = cluster_true,
  MCMC     = clustering_matrix[best_ids["MCMC"], ],
  PostMean = clustering_matrix[best_ids["PostMean"], ],
  MLE      = clustering_matrix[best_ids["MLE"], ],
  Binder   = clustering_matrix[best_ids["Binder"], ]
)

# --- ARI vs truth ---
ari_df <- tibble(
  Method = names(best_clusterings),
  ARI = sapply(best_clusterings, function(cl) adjustedRandIndex(cluster_true, cl))
)

# --- Long plotting data ---
plot_long_2d <- bind_rows(
  lapply(names(best_clusterings), function(m) {
    tibble(
      x1 = y[,1],
      x2 = y[,2],
      Cluster = factor(best_clusterings[[m]]),
      Method = m
    )
  })
)

# --- ARI label positions (top-right of each facet) ---
label_pos <- plot_long_2d %>%
  group_by(Method) %>%
  summarise(
    x_pos = max(x1),
    y_pos = max(x2),
    .groups = "drop"
  )

ari_df <- ari_df %>%
  left_join(label_pos, by = "Method") %>%
  mutate(label = paste0("ARI: ", round(ARI, 3)))

# --- Plot ---
ggplot(plot_long_2d, aes(x = x1, y = x2, color = Cluster)) +
  geom_point(alpha = 0.75, size = 2.5) +
  geom_text(
    data = ari_df,
    aes(x = x_pos, y = y_pos, label = label),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 4,
    color = "black"
  ) +
  facet_wrap(~ Method, ncol = 2) +
  theme_minimal() +
  labs(
    x = expression(y[1]),
    y = expression(y[2]),
    color = "Cluster",
    title = "Best Clusterings per Method (2D) with ARI vs Truth"
  )


```

The MLE solution seems to be doing very good!

### Correlation with binder

```{r}

# Add binder_ests to results_df (if not already)
results_df <- results_df %>%
  mutate(binder_ests = binder_ests)

# Gather distance columns into long format for faceting
# Replace with your actual distance column names
results_long <- results_df %>%
  select(binder_ests, dist_mcmc, dist_postmean, dist_mle) %>%
  pivot_longer(
    cols = starts_with("dist_"),
    names_to = "method",
    values_to = "distance"
  ) %>%
  mutate(
    method = factor(method,
                    levels = c("dist_mcmc", "dist_postmean", "dist_mle"),
                    labels = c("MCMC posterior", "Posterior Mean", "MLE"))
  )


# Plot with facets by method
p <- ggplot(results_long, aes(x = distance, y = binder_ests)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  facet_wrap(~ method, scales = "free", ncol = 1) +
  labs(
    x = "Distance",
    y = "Binder Loss",
    title = "Comparison of Distance Metrics vs Binder Loss"
  ) +
  theme_minimal()

print(p)

```


Weirdly not even the MLE solution seems to correlate with the binder estimates.


# 3 D

```{r}

# -----------------------------
# Simulation parameters
# -----------------------------
N <- 100
K <- 3
D = 3
alpha_0 <- 1

set.seed(2)
  
simdata <- generate_mixture_data(
    N,
    K,
    dim = D,
    alpha = alpha_0,
    mu_true = NULL,      
    Sigma_true = NULL 
)

y <- simdata$data
cluster_true <- simdata$cluster_true
weights_true <- simdata$weights_true
Sigma_true <- simdata$Sigma_true

library(plotly)

stopifnot(ncol(y) == 3)

plot_ly(
  x = y[, 1],
  y = y[, 2],
  z = y[, 3],
  color = factor(cluster_true),
  colors = "Set1",
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 5)
) %>%
  layout(
    scene = list(
      xaxis = list(title = "y1"),
      yaxis = list(title = "y2"),
      zaxis = list(title = "y3")
    ),
    legend = list(title = list(text = "True cluster"))
  )



```


```{r}
# -----------------------------
# Simulation parameters
# -----------------------------
N <- 100
K <- 3
D = 3
alpha_0 <- 1

set.seed(2)
  
simdata <- generate_mixture_data(
    N,
    K,
    dim = D,
    alpha = alpha_0,
    mu_true = NULL,      
    Sigma_true = NULL 
)

y <- simdata$data
cluster_true <- simdata$cluster_true
weights_true <- simdata$weights_true
Sigma_true <- simdata$Sigma_true

# -----------------------------
# Prior hyperparameters
# -----------------------------
alpha <- 1
kappa_0 <- 0.1
Lam_0 <- diag(D) 
mu_0 <- rep(0, D)
nu_0 <- D + 2

# -----------------------------
# MCMC settings
# -----------------------------
n_iter <- 1e3
burn_in <- n_iter / 4
thin <- 10

# -----------------------------
# Define ANTMAN priors & parameters
# -----------------------------
mixture_mvn_params <- AntMAN::AM_mix_hyperparams_multinorm(
 mu0 = mu_0, ka0 = kappa_0, nu0 = nu_0, Lam0 = Lam_0
)

mcmc_params <- AntMAN::AM_mcmc_parameters(
  niter = n_iter, burnin = burn_in, thin = thin, verbose = 1
)

components_prior <- AntMAN::AM_mix_components_prior_pois(Lambda = K)

# weights_prior <- AntMAN::AM_mix_weights_prior_gamma(init = 2, a = 1, b = 1)

# -----------------------------
# Run MCMC and cluster
# -----------------------------
mix_post_draws <- AntMAN::AM_mcmc_fit(
  y = y,
  mix_kernel_hyperparams = mixture_mvn_params,
  mix_components_prior = components_prior,
  # mix_weight_prior = weights_prior,
  mcmc_parameters = mcmc_params
)

eam <- AM_clustering(mix_post_draws)  # M*n matrix

```

Helper function for the following analysis:

```{r}

compute_distances <- function(
  data,
  clustering_matrix,
  param_method = c("mcmc", "postmean", "mle"),
  prior_list = NULL,
  posterior_params = NULL,
  distance_method = "KS"
) {
  
  param_method <- match.arg(param_method)
  n_candidates <- nrow(clustering_matrix)
  dist <- numeric(n_candidates)
  
  for (i in seq_len(n_candidates)) {
    params <- get_mixture_params(
      clustering       = clustering_matrix[i, ],
      data             = data,
      param_method     = param_method,
      prior_list       = prior_list,
      posterior_params = posterior_params,
      candidate_id     = i
    )
    
    dist[i] <- compute_distance(
      data,
      params$weight,
      params$mean,
      params$var,
      method = distance_method
    )
  }
  
  dist
}



```

```{r}

prior_list <- list(
  alpha_0 = 1,
  kappa_0 = 1,
  mu_0    = 0,
  a_0     = 2,
  b_0     = 1,
  nu_0    = 5,
  S_0     = diag(D)
)

data = y
clustering_matrix <- eam + 1
posterior_params <- prepare_AM_posterior_params(mix_post_draws)

dist_mcmc <- compute_distances(
  data = data,
  clustering_matrix = clustering_matrix,
  param_method = "mcmc",
  posterior_params = posterior_params,
  distance_method = "Wasserstein"
)

dist_postmean <- compute_distances(
  data = data,
  clustering_matrix = clustering_matrix,
  param_method = "postmean",
  prior_list = prior_list,
  distance_method = "Wasserstein"
)

dist_mle <- compute_distances(
  data = data,
  clustering_matrix = clustering_matrix,
  param_method = "mle",
  distance_method = "Wasserstein"
)

binder_ests <- rep(NA, n_candidates)
vi_ests <- rep(NA, n_candidates)

for(i in 1:nrow(eam)) {
binder_ests[i] <- salso::binder(cluster_true,  clustering_matrix[i, ])
vi_ests[i] <- salso::VI(cluster_true,  clustering_matrix[i, ])
}


results_df <- data.frame(
  candidate     = seq_len(nrow(clustering_matrix)),
  dist_mcmc     = dist_mcmc,
  dist_postmean = dist_postmean,
  dist_mle      = dist_mle
)

print(results_df)


results_df$binder_ests <- binder_ests 
results_df$vi_ests <- vi_ests 

```



```{r}

library(mclust)
library(dplyr)
library(ggplot2)

# --- Select best clusterings ---
best_ids <- c(
  Truth     = NA,
  MCMC      = which.min(results_df$dist_mcmc),
  PostMean  = which.min(results_df$dist_postmean),
  MLE       = which.min(results_df$dist_mle),
  Binder    = which.min(results_df$binder_ests),
  VI    = which.min(results_df$vi_ests)
)

best_clusterings <- list(
  Truth    = cluster_true,
  MCMC     = clustering_matrix[best_ids["MCMC"], ],
  PostMean = clustering_matrix[best_ids["PostMean"], ],
  MLE      = clustering_matrix[best_ids["MLE"], ],
  Binder   = clustering_matrix[best_ids["Binder"], ],
  VI   = clustering_matrix[best_ids["VI"], ]
)

# --- ARI vs truth ---
ari_df <- tibble(
  Method = names(best_clusterings),
  ARI = sapply(best_clusterings, function(cl) adjustedRandIndex(cluster_true, cl))
)

# --- Long plotting data ---
plot_long_2d <- bind_rows(
  lapply(names(best_clusterings), function(m) {
    tibble(
      x1 = y[,1],
      x2 = y[,2],
      Cluster = factor(best_clusterings[[m]]),
      Method = m
    )
  })
)

# --- ARI label positions (top-right of each facet) ---
label_pos <- plot_long_2d %>%
  group_by(Method) %>%
  summarise(
    x_pos = max(x1),
    y_pos = max(x2),
    .groups = "drop"
  )

ari_df <- ari_df %>%
  left_join(label_pos, by = "Method") %>%
  mutate(label = paste0("ARI: ", round(ARI, 3)))

# --- Plot ---
ggplot(plot_long_2d, aes(x = x1, y = x2, color = Cluster)) +
  geom_point(alpha = 0.75, size = 2.5) +
  geom_text(
    data = ari_df,
    aes(x = x_pos, y = y_pos, label = label),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 4,
    color = "black"
  ) +
  facet_wrap(~ Method, ncol = 2) +
  theme_minimal() +
  labs(
    x = expression(y[1]),
    y = expression(y[2]),
    color = "Cluster",
    title = "Best Clusterings per Method (2D) with ARI vs Truth"
  )


```

Still seems to work!

3D visualization

```{r}

library(plotly)
library(mclust)
library(dplyr)

stopifnot(ncol(y) == 3)

# -----------------------------
# Select best clusterings
# -----------------------------
best_ids <- c(
  Truth    = NA,
  MCMC     = which.min(results_df$dist_mcmc),
  PostMean = which.min(results_df$dist_postmean),
  MLE      = which.min(results_df$dist_mle),
  Binder   = which.min(results_df$binder_ests),
  VI       = which.min(results_df$vi_ests)
)

best_clusterings <- list(
  Truth    = cluster_true,
  MCMC     = clustering_matrix[best_ids["MCMC"], ],
  PostMean = clustering_matrix[best_ids["PostMean"], ],
  MLE      = clustering_matrix[best_ids["MLE"], ],
  Binder   = clustering_matrix[best_ids["Binder"], ],
  VI       = clustering_matrix[best_ids["VI"], ]
)

# -----------------------------
# ARI vs truth
# -----------------------------
ari_vals <- sapply(
  best_clusterings,
  function(cl) adjustedRandIndex(cluster_true, cl)
)

# -----------------------------
# Build 3D plots per method
# -----------------------------
plots_3d <- lapply(names(best_clusterings), function(m) {
  
  plot_ly(
    x = y[,1],
    y = y[,2],
    z = y[,3],
    color = factor(best_clusterings[[m]]),
    colors = "Set1",
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 4)
  ) %>%
    layout(
      title = paste0(
        m,
        " | ARI = ",
        round(ari_vals[m], 3)
      ),
      scene = list(
        xaxis = list(title = expression(y[1])),
        yaxis = list(title = expression(y[2])),
        zaxis = list(title = expression(y[3]))
      ),
      legend = list(title = list(text = "Cluster"))
    )
})

plots_3d[[1]]

# # -----------------------------
# # Combine panels
# # -----------------------------
# subplot(
#   plots_3d,
#   nrows = 3,
#   margin = 0.04,
#   shareX = TRUE,
#   shareY = TRUE,
#   titleX = TRUE,
#   titleY = TRUE
# )


```




